---
title: Stochastic methods for Bayesian inference
author: Joshua French
date: ''
# format:
#   revealjs:
#     slide-level: 1
# self-contained: true
# execute:
#   echo: true
jupyter:
  kernelspec:
    name: "ir"
    language: "R"
    display_name: "R"
    path: ""
format: ipynb
execute:
  output: false
title-block-banner: true
wrap: 'none'
---

An interactive version of this content is available as a Colab notebook, which can be accessed by scanning or clicking the QR code below. 

<a href="https://colab.research.google.com/github/jfrench/BayesianStatistics/blob/master/ch03c-stochastic.ipynb">
  <img src="https://raw.githubusercontent.com/jfrench/BayesianStatistics/refs/heads/master/qr_codes/stochastic.svg" width="300" height="300">
    </a>

---

# Stochastic methods

---

## Rejection sampling

- Rejection sampling can be used to draw samples from a target density $p(\theta \mid y)$ or its unnormalized version $q(\theta \mid y)$.
- Rejection sampling assumes there is a *proposal distribution* $G$, with density function $g(\theta)$, such that:
    -   We can easily draw a realization from $G$.
    -   If $p(\theta = \theta^{\star} \mid y) > 0$, then $g(\theta = \theta^{\star} ) > 0$.
    -   The *importance ratio* $p(\theta \mid y) / g(\theta)$ has known bound $M$.
    -   i.e., $p(\theta\mid y) / g(\theta) \leq M$ for all $\theta$.

---

Rejection sampling draws $B$ observations from $p(\theta\mid y)$ using the following algorithm:

1.  Draw $\theta^{\star}\sim G$.
2.  Accept $\theta^{\star}$ with probability
  $$
  \frac{p(\theta^{\star}\mid y)}{M g(\theta^{\star})}.
  $$
  Otherwise, discard it.
3.  Repeat steps 1 and 2 until $B$ candidates have been accepted.

---

An alternative approach to decide whether we accept the candidate value  $\theta^{\star}$  is:

- Draw $u^{\star}\sim U(0, M g(\theta^{\star} ) )$.
- Retain $\theta^{\star}$ if $u^{\star} \leq p(\theta^{\star} \mid y)$.

---

Comments about rejection sampling:

-   The unnormalized density, $q(\theta \mid y)$, is almost always used instead of the true density, $p(\theta \mid y)$
-   $p(\theta \mid y) \leq M g(\theta \mid y)$ ensures the acceptance probability is no larger than 1.
-   $M$ should be chosen so that $M g(\theta \mid y)$ is as close as possible to $p(\theta\mid y)$.

---

## Rejection sampling example 1

-   Suppose the (unnormalized) target density is $q(\theta \mid y) = \theta(1-\theta)I_{(0,1)}(\theta).$
-   $q(\theta\mid y)$ is bounded by a $U(0, 1)$ density, i.e., $g(\theta)=I_{(0,1)}(\theta)$.
-   $\text{argmax}_{\theta}q(\theta \mid y) = 0.5$ with $q(0.5\mid y) = 0.25$.
-   Pick $M = 0.25$.

---

We define the unnormalized target density. 

```{r}
qtarget <- function(theta) {
  theta * (1 - theta) * (theta > 0) * (theta < 1)
}
```

We define the bounding function, $g(\theta)M$.

```{r}
# define bounding function
gM = function(theta) {
  rep(.25, length(theta))
}
```

---

We visualize the target density and bounding function.

```{r}
theta <- seq(0, 1, len = 10000)
par(mar = c(4, 4, 0.4, 0.4)) # edit margins
plot(theta, qtarget(theta), type = "l", xlab = expression(theta),
     ylab = expression(q(theta*"|"*y)))
lines(theta, gM(theta), col = "blue")
```

---

Consider drawing an observation from $p(\theta \mid y)$:

-   We draw a value, $\theta^*$, from the $U(0,1)$ proposal distribution.
    -   Suppose this value is $\theta^*=0.6$.
-   We need to decide whether we retain or reject $\theta^*$.
    -   Note that $M g(\theta)$ evaluated at $0.6$ is $0.25 \times 1 = 0.25$.
    -   Note that $q(0.6 \mid y) = 0.6(1-0.6) = 0.24$.
-   Draw $u^{\star}$ from a $U(0, 0.25)$ density.
    - Let's say $u^{\star} = 0.066$.
    - Since $u^{\star}<q(\theta\mid y) = 0.24$, we accept it as a sample from $q(\theta\mid y)$.

---

We visualize this.

```{r}
#| output: false
thetastar1 <- 0.6
u1 <- 0.066
accept1 <- (u1 <= qtarget(thetastar1))

# plot q and bounding function
theta = seq(0, 1, len = 1000)
plot(theta, qtarget(theta), type = "l", xlab = expression(theta),
     ylab = expression(q(theta*"|"*y)))
lines(theta, gM(theta), col = "blue")
abline(v = thetastar1, col = "grey")
points(thetastar1, u1, pch = ifelse(accept1, 20, 1))
```

---

```{r}
#| echo: false
thetastar1 <- 0.6
u1 <- 0.066
accept1 <- (u1 <= qtarget(thetastar1))

# plot q and bounding function
theta <- seq(0, 1, len = 1000)
plot(theta, qtarget(theta), type = "l", xlab = expression(theta),
     ylab = expression(q(theta*"|"*y)))
lines(theta, gM(theta), col = "blue")
abline(v = thetastar1, col = "grey")
points(thetastar1, u1, pch = ifelse(accept1, 20, 1))
```

---

Consider drawing another observation from $p(\theta \mid y)$:

-   We draw a value $\theta^*$ from the $U(0,1)$ distribution.
    -   Suppose this value is $\theta^*=0.185$.
-   We draw $u^{\star}=0.176$ from $U(0, Mg(\theta^*))=U(0, 0.25g(0.185))=U(0, 0.25)$.
-   $u^{\star}=0.176 > 0.151 = q(\theta^{\star}|y) = q(0.185|y)$, so we reject the proposed value of $\theta^{\star}$.


---

We visualize this.

```{r}
#| eval: false
thetastar2 <- 0.185
u2 <- 0.176
accept2 <- (u2 <= qtarget(thetastar2))
# plot q and bounding function
theta <- seq(0, 1, len = 1000)
plot(theta, qtarget(theta), type = "l", xlab = expression(theta),
     ylab = expression(q(theta*"|"*y)))
lines(theta, gM(theta), col = "blue")
abline(v = thetastar2, col = "grey")
points(thetastar2, u2, pch = ifelse(accept2, 20, 1))
```

---

```{r}
#| echo: false
thetastar2 <- 0.185
u2 <- 0.176
accept2 <- (u2 <= qtarget(thetastar2))
# plot q and bounding function
theta <- seq(0, 1, len = 1000)
plot(theta, qtarget(theta), type = "l", xlab = expression(theta),
     ylab = expression(q(theta*"|"*y)))
lines(theta, gM(theta), col = "blue")
abline(v = thetastar2, col = "grey")
points(thetastar2, u2, pch = ifelse(accept2, 20, 1))
```

---

- We run a rejection sampling algorithm to draw observations
from the target distribution.
- We plot closed circles for each $(\theta^*, u^*)$ that's accepted.
- We plot a closed circle for each $(\theta^*, u^*)$ that's rejected.

```{r}
#| eval: false
set.seed(8)
plot(theta, qtarget(theta), type = "l", xlab = expression(theta),
     ylab = expression(q(theta*"|"*y)))
lines(theta, gM(theta), col = "blue")
for(i in 1:25) {
  thetastar <- runif(1)
  u <- runif(1, 0, gM(thetastar))
  accept <- (u <= qtarget(thetastar))
  points(thetastar, u, pch = ifelse(accept, 20, 1))
}
title("Accepted (black dot) vs Rejected (open circle) comparison")
```

---

```{r}
#| echo: false
set.seed(8)
plot(theta, qtarget(theta), type = "l", xlab = expression(theta),
     ylab = expression(q(theta*"|"*y)))
lines(theta, gM(theta), col = "blue")
for(i in 1:25) {
  thetastar <- runif(1)
  u <- runif(1, 0, gM(thetastar))
  accept <- (u <= qtarget(thetastar))
  points(thetastar, u, pch = ifelse(accept, 20, 1))
}
title("Accepted (black dot) vs Rejected (open circle) comparison")
```

---

## Rejection sampling example 2

We illustrate rejection sampling by trying to draw realizations from an unnormalized Beta(3, 3) density. 

**Target density**

$p(\theta | y) \propto \theta^2  (1 - \theta)^2 I(0<\theta<1)= q(\theta).$

**Proposal distribution**

Since $q(\theta)$ has finite support, our proposal distribution can be $G\sim U(0,1)$. 

- We need to find the bounding constant $M$ for our bounding function.

---

- Taking the derivative of $q(\theta)$ with respect to $\theta$,
we can determine that the function has a maximum at $\theta = 1/2$.
- The maximum is $q(1/2) = 0.0625$.
- Thus, $q(\theta) <= g(\theta)M$ with $M = 0.0625$.

We define the unnormalized target density.

```{r}
qtarget <- function(theta) {
	theta^2*(1 - theta)^2 * (theta > 0) * (theta < 1)
}
```

---

We define the bounding function.

```{r}
gM <- function(theta) {
	rep(.0625, length(theta))
}
```

We compare the target density and bounding function.

```{r}
#| eval: false
theta <- seq(0, 1, len = 1000)
plot(theta, qtarget(theta), type = "l", xlab = expression(theta),
     ylab = expression(q(theta)))
lines(theta, gM(theta), col = "blue")
abline(v = 1/2)
```

---

```{r}
#| echo: false
theta <- seq(0, 1, len = 1000)
plot(theta, qtarget(theta), type = "l", xlab = expression(theta),
     ylab = expression(q(theta)))
lines(theta, gM(theta), col = "blue")
abline(v = 1/2)
```

---

We implement our rejection sampling algorithm.

```{r}
B <- 10000 # number of samples to keep
mytheta <- numeric(B) # vector to store kept samples
set.seed(57) # reproducibility
i <- 0 # the samples accepted
while (i < B) {
  x <- runif(1) # sample from G
  # accept x with probability q(x)/gM(x)
	if (runif(1) <= qtarget(x)/gM(x)) {
		i <- i + 1
		mytheta[i] <- x
	}
}
```

---

We compare our empirical density to the true density.

```{r}
#| eval: false
dapprox <- density(mytheta)
dtruth <- dbeta(theta, 3, 3)

plot(theta, dtruth, xlab = expression(theta), ylab = "density",
     main = "", type = "l", col = "blue")
lines(dapprox, col = "orange")
legend("topleft", legend = c("approximation", "truth"),
       col = c("orange", "blue"), lwd = c(1, 1))
```

---
  
```{r}
#| echo: false
dapprox <- density(mytheta)
dtruth <- dbeta(theta, 3, 3)

plot(theta, dtruth, xlab = expression(theta), ylab = "density", main = "", type = "l", col = "blue")
lines(dapprox, col = "orange")
legend("topleft", legend = c("approximation", "truth"),
       col = c("orange", "blue"), lwd = c(1, 1))
```

---

## Rejection sampling example 3

We repeat the previous example with a more efficient proposal distribution.

We know that our target density is symmetric around 0.5, we choose to bound it by a normal distribution centered around 0.5.

**Proposal distribution**

$G\sim N(0.5, 0.25^2)$.

We choose $M=0.04$ for our bounding function (chosen by trial and error).

---

We define bounding function.

```{r}
gM <- function(theta) {
	0.04 * dnorm(theta, mean = 0.5, sd = 0.25)
}
```

We compare the target density to the bounding function.

```{r}
#| eval: false
theta <- seq(0, 1, len = 1000)
plot(theta, qtarget(theta), type = "l", xlab = expression(theta),
     ylab = expression(q(theta)))
lines(theta, gM(theta), col = "blue")
abline(v = 1/2)
```

---

```{r}
#| echo: false
theta <- seq(0, 1, len = 1000)
plot(theta, qtarget(theta), type = "l", xlab = expression(theta),
     ylab = expression(q(theta)))
lines(theta, gM(theta), col = "blue")
abline(v = 1/2)
```

---

We implement our rejection sampling algorithm.

```{r}
set.seed(71)
B <- 100000
mytheta = numeric(B)
i <- 0 # number of candidates retained
while (i < B) {
  x <- rnorm(1, mean = 0.5, sd = 0.25) # sample from g distribution
  # accept x with probability q(x)/gM(x)
	if (runif(1) <= qtarget(x)/gM(x)) {
		i <- i + 1
		mytheta[i] <- x
	}
}
```

---

We compare our empirical density to the true density.

```{r}
#| eval: false
dmytheta <- density(mytheta)
dtruth <- dbeta(theta, 3, 3)
plot(theta, dtruth, col = "blue", type = "l", ylab = "density")
lines(dmytheta, col = "orange")
legend("topleft", legend = c("approximation", "truth"),
       col = c("orange", "blue"), lwd = c(1, 1))
```

---

```{r}
#| echo: false
dmytheta <- density(mytheta)
dtruth <- dbeta(theta, 3, 3)
plot(theta, dtruth, col = "blue", type = "l", ylab = "density")
lines(dmytheta, col = "orange")
legend("topleft", legend = c("approximation", "truth"),
       col = c("orange", "blue"), lwd = c(1, 1))
```


---

## Rejection sampling example 4

We want to sample from a folded $N(0,1)$ distribution.

- If $y\sim N(0,1)$, then $|y|$ is a folded $N(0,1)$ distribution.

**Target density**

$$
q(\theta) = \exp(-\theta^2/2) I_{[0,\infty)}(\theta)
$$

**Proposal distribution**

$G\sim \text{Exp}(1)$

---

How do we choose a bounding constant?

The optimal solution is find a single intersecting point between the folded normal and our envelope (at the inflection point of the folded normal), which is at $\theta = 1$.

- $\frac{dq(\theta)}{d\theta} = -\exp(-\theta^2/2)\theta$.
- $\frac{d^2q(\theta)}{d\theta^2} = \exp(-\theta^2/2)(\theta^2-1)$.
- Setting $g(\theta)M = q(\theta)$ and solving for $M$ when
$\theta = 1$ results in the solution $M = \exp(1/2)$.

---

We define the target density.

```{r}
qtarget <- function(theta) {
	exp(-theta^2/2) * (theta >= 0)
}
```

We define the bounding function.

```{r}
gM <- function(theta) {
	dexp(theta) * exp(1/2)
}
```

We compare the target density to the bounding function.

```{r}
#| eval: false
theta <- seq(0, 5, len = 1000)
plot(theta, qtarget(theta), ylim = c(0, qtarget(0.001)),
     type = "l", ylab = "qtarget", xlab = expression(theta))
lines(theta, gM(theta), type = "l", col = "blue")
legend("topright", legend = c("qtarget", "gM"),
       lty = 1, col = c("black", "blue"))
```

---

```{r}
#| echo: false
theta <- seq(0, 5, len = 1000)
plot(theta, qtarget(theta), ylim = c(0, qtarget(0.001)),
     type = "l", ylab = "qtarget", xlab = expression(theta))
lines(theta, gM(theta), type = "l", col = "blue")
legend("topright", legend = c("qtarget", "gM"),
       lty = 1, col = c("black", "blue"))
```

---

We implement our rejection sampling algorithm.

```{r}
B <- 10000 # number of retained samples desired
mytheta <- numeric(B) # vector to store samples
i = 0 # number of retained samples
while (i < B) {
	x <- rexp(1) # draw a value from proposal distribution
	# accept the value with probability based on the importance
	# ratio
	if (runif(1) <= qtarget(x)/gM(x)) {
		i <- i + 1 # increment i if value retained
		mytheta[i] <- x # store value
	}
}
```

---

We compare our empirical density to the true density.

```{r}
#| eval: false
# true density
dens <-  function(theta) { sqrt(2/pi)*exp(-theta^2/2) }
dtruth <- dens(theta)
# approximate density
dapprox <- density(mytheta, from = 0, to = 5, cut = 0)
plot(theta, dtruth, ylab = "density", main = "", type = "l",
     col = "blue")
lines(dapprox, col = "orange")
legend("topright", legend = c("approximation", "truth"),
	lwd = c(1, 1), col = c("orange", "blue"))
```

---

```{r}
#| echo: false
# true density
dens <-  function(theta) { sqrt(2/pi)*exp(-theta^2/2) }
dtruth <- dens(theta)
# approximate density
dapprox <- density(mytheta, from = 0, to = 5, cut = 0)
plot(theta, dtruth, ylab = "density", main = "", type = "l",
     col = "blue")
lines(dapprox, col = "orange")
legend("topright", legend = c("approximation", "truth"),
	lwd = c(1, 1), col = c("orange", "blue"))
```

--- 

Why didn't that work properly?

We try a different approach using a histogram of the retained samples.

```{r}
#| eval: false
# plot with a probability histogram instead
hist(mytheta, freq = FALSE, breaks = 100, main = "")
lines(theta, truth, col = "blue")
legend("topright", legend = c("approximation", "truth"),
	lwd = c(1, 1), col = c("black", "blue"))
```

---

```{r}
#| echo: false
# plot with a probability histogram instead
hist(mytheta, freq = FALSE, breaks = 100, main = "")
lines(theta, dtruth, col = "blue")
legend("topright", legend = c("approximation", "truth"),
	lwd = c(1, 1), col = c("black", "blue"))
```

---

## Rejection sampling example 5

- We chose a proposal distribution analytically in the last example.
- We probably wouldn't do that unless we were forced to.
- Let's try something more realistic.

**Proposal distribution**

$G \sim \chi^2_2$ with bounding constant $M=2.4$ (chosen by trial and error).

--- 

We create the bounding function.

```{r}
gM <- function(theta) {
	2.4 * dchisq(theta, df = 2)
}
```

We compare the target density and bounding function.

```{r}
#| eval: false
theta <- seq(0, 5, len = 10000)
plot(theta, gM(theta), type = "l", col = "blue",
     xlab = expression(theta), ylab = "", ylim = c(0, 1.25))
lines(theta, qtarget(theta))
legend("topright", legend = c("q", "gM"), lty = 1,
       col = c("black", "blue"))
```

---

```{r}
#| echo: false
theta <- seq(0, 5, len = 10000)
plot(theta, gM(theta), type = "l", col = "blue",
     xlab = expression(theta), ylab = "", ylim = c(0, 1.25))
lines(theta, qtarget(theta))
legend("topright", legend = c("q", "gM"), lty = 1,
       col = c("black", "blue"))
```

---

We implement our rejection sampling algorithm.

```{r}
B = 10000
mytheta = numeric(B)
i = 0
while (i < B) {
	x = rchisq(1, df = 2)
	if (runif(1) <= qtarget(x)/gM(x)) {
		i = i + 1
		mytheta[i] = x
	}
}
```

---

We compare our empirical density to the true density.

```{r}
#| eval: false
hist(mytheta, freq = FALSE, breaks = 100, main = "")
lines(theta, dens(theta), col = "blue")
legend("topright", legend = c("approximation", "truth"),
       lwd = c(1, 1), col = c("black", "blue"))
```

---

```{r}
#| echo: false
hist(mytheta, freq = FALSE, breaks = 100, main = "")
lines(theta, dens(theta), col = "blue")
legend("topright", legend = c("approximation", "truth"),
       lwd = c(1, 1), col = c("black", "blue"))
```