---
title: Computational methods for Bayesian inference
author: Joshua French
date: ''
# format: html
format: ipynb
jupyter:
  kernelspec:
    display_name: R
    language: R
    name: ir
execute:
  output: false
self-contained: true
title-block-banner: true
wrap: 'none'
---

To open this information in an interactive Colab notebook, click the Open in Colab graphic below.

<a href="https://colab.research.google.com/github/jfrench/BayesianStatistics/blob/master/lectures/Math7393-03d-Computational-methods-for-Bayesian-inference-Gibbs.ipynb">
   <img src="https://colab.research.google.com/assets/colab-badge.svg">
</a>

We will need certain packages to be available for this notebook. The code below will check if they are installed and install them if necessary.

```{r}
# see if remotes package is available
if (!require("remotes", quietly = TRUE)) {
  # if not, then install package
  install.packages(
    "remotes",
    repos = "https://cran.rstudio.com/"
    )
}
```

```{r}
# see if bayesutils package is available
if (!require("bayesutils", quietly = TRUE)) {
  # if not, then install package
  remotes::install_github("jfrench/bayesutils")
  library(bayesutils)
}
```

```{r}
# see if autoimage package is available
if (!require("autoimage", quietly = TRUE)) {
  # if not, then install package
  install.packages("autoimage")
  library(autoimage,
          repos = "https://cran.rstudio.com/",
          quietly = TRUE)
}
```

# Introduction to MCMC Methods

A *stochastic process* is a family of random variables $\{\theta^{(t)}\}_{t\in T}$, where $T$ is a subset of $[0, \infty)$.
  -   $\theta^{(t)}$ denotes the stochastic process at time $t$.
  -   The value $\theta^{(t)}$ takes at time $t$ is known as the *state* of the process at time $t$.
  -   We only consider discrete-time stochastic processes with $T=\{0,1,2,…\}$.

The *state space* of a stochastic process is the set of all possible values the process takes.

A *Markov chain* is a stochastic process $\{\theta^{(0)}, \theta^{(1)}, \ldots, \theta^{(B)}\}$ such that $$p(\theta^{(t+1)} | \theta^{(t)}, \theta^{(t-1)}, \ldots, \theta^{(0)}) = p(\theta^{(t+1)} | \theta^{(t)}).$$
  -   $p(\theta^{(t+1)} | \theta^{(t)})$ is often called the *transition distribution*.

Markov chain Monte Carlo (MCMC) methods attempt to draw samples from a target distribution when sampling directly from the target distribution is impossible or computationally expensive.
  - The samples form a Markov chain.
  - MCMC methods are constructed so that the (limiting) distribution of the Markov chain converges to the target distribution.
      -   The distribution producing the samples at each step of the Markov chain becomes more like the target distribution at each step of the chain.

A Markov chain converges to a *stationary distribution* when the Markov chain is irreducible, aperiodic, and positive recurrent.
    - Irreducible: any state $j$ can be reached from any state $i$ in a finite number of steps.
    - Positive recurrent: with probability 1, the chain starting in state i returns to state i in a finite number of steps for any starting state.
    - Aperiodic: the chain does not travel between states in a specific pattern or number of steps.
    - Combined: for any state $i$ and $j$, we can travel from $i$ to $j$ in a finite number of steps, and then we will travel back to $i$ with probability 1, but not in a specific pattern or number of steps.
    -   This ensures that we adequately explore the entire support of our target distribution.

An effective MCMC method has two desirable properties:

1.  It must be easy to draw from the $p(\theta^{(t+1)}|\theta^{(t)})$.
2.  The stationary distribution of the Markov chain must match our target distribution.

To construct a Markov chain using an MCMC algorithm:

-   Specify an initial value, $\theta^{(0)}$, for the chain.
    -   Often this is a "likely" value of $\theta$.
-   For $t = 1,2, \ldots, B$, draw samples from $p(\theta^{(t+1)}|\theta^{(t)})$ until $B$ is large enough that the stationary distribution is reached.
-   After the chain has been implemented, check the convergence of the simulated sequence (or at least verify that there isn't clear evidence that the chain hasn't converged).
    -   This is VERY important.

Concluding thoughts:

-   The *Gibbs* and *Metropolis-Hastings* algorithms are the most well-known MCMC algorithms.
    -   They are designed so that the stationary distribution matches the target distribution.
-   MCMC methods are most popularly used in Bayesian statistics.
    -   In Bayesian statistics, the target distribution is generally the posterior distribution, $p(\theta | y)$.

# Gibbs sampling

-   The *Gibbs sampling* algorithm (or *Gibbs sampler*) is the most popular Markov chain Monte Carlo (MCMC) method.
-   The goal of the Gibbs sampler is to draw samples from a target distribution.
-   The Gibbs sampling algorithm produces samples from the target distribution by successively drawing samples from the full conditional distributions of the target distribution.
    -   The *full conditional distribution* of a random vector is the distribution of the random vector conditional on all the other random variables in the joint (i.e., target) distribution.

Some notation and terminology:

-   Let $\theta$ be the vector of random variables comprising all variables of the target distribution, $p(\theta | y)$.
-   Partition $\theta$ into $d$ components, i.e., $\theta = (\theta_1, \theta_2, \ldots, \theta_d)$.
    -   Each component could be a (single) random variable or a random vector.
-   $\theta_{-j}$ denotes the vector containing all components in $\theta$ except $\theta_j$, i.e., $$\theta_{-j} = (\theta_1, \theta_2, \ldots, \theta_{j-1}, \theta_{j+1}, \ldots, \theta_d).$$
-   The *full conditional distribution* of $\theta_j$, denoted $p(\theta_j | \theta_{-j}, y)$, is the distribution of component $\theta_j$ conditional on knowing the value of all other components, $\theta_{-j}$, and the data, $y$.
-   After we have drawn samples from the full conditional distribution of each component (i.e., drawn samples from $p(\theta_{j} | \theta_{-j}^{(t-1)}, y)$ for $j=1,2,\ldots,d$) we have completed a *cycle*.

More notation:

-   $\theta_j^{(t)}$ denotes the sampled value of $\theta_j$ in cycle $t$ and $\theta^{(t)}$ the vector of all values sampled in cycle $t$.
-   $\theta_{-j}^{(t-1)}$ denotes the most current value of all $d$ components of $\theta$ **except** $\theta_j$, i.e., $$\theta_{-j}^{(t-1)} = (\theta_1^{(t)}, \theta_2^{(t)}, \ldots, \theta_{j-1}^{(t)}, \theta_{j+1}^{(t-1)}, \ldots, \theta_{d}^{(t-1)}).$$
-   $p(\theta_j | \theta_{-j}^{(t-1)}, y)$ denotes the full conditional distribution of $\theta_j$ conditional on $\theta_{-j}$ being fixed at $\theta_{-j}^{(t-1)}$ and the data being fixed at $y$.

The basic Gibbs sampling algorithm is:

1.  Choose starting values for all of your components, i.e., $\theta^{(0)} = (\theta^{(0)}_1, \theta^{(0)}_2, \ldots, \theta^{(0)}_d)$.
2.  Set $t = 1$.
3.  Draw $\theta_j^{(t)}$ from the full conditional distribution $p(\theta_j | \theta_{-j}^{(t-1)}, y)$ for $j = 1, 2, \ldots, d$.
4.  Increment $t$.
5.  Repeat steps 3 and 4 until convergence.

# Example: Gibbs sampler (bivariate normal)

This example appears in Bayesian Data Analysis, 3rd edition, by Gelman et al. (2013).

*Data distribution* 

$y | \theta \sim N(\theta, \Sigma)$ is a bivariate normal distribution with unknown mean 
$\theta = (\theta_1, \theta_2)$ and known covariance matrix 

$$
\Sigma =
\begin{bmatrix}
 1 & \rho \\
\rho & 1
\end{bmatrix}.
$$

*Prior distribution* 

The prior for $\theta$ is an improper uniform over the real line, i.e., $p(\theta_1,\theta_2)\propto1$.

(Why? Because it makes the example easier!)

*Posterior distribution* 

Assuming we observe a single observation $y=(y_1,y_2)$, $$\theta | y \sim N(y, \Sigma).$$

Let's derive the full conditional distributions for $\theta_1$ and $\theta_2$.

<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  

*Full conditional distributions*

$\theta_1 | \theta_2, y \sim N(y_1 + \rho(\theta_2-y_2),1-\rho^2)$

$\theta_2 | \theta_1, y \sim N(y_2 + \rho(\theta_1-y_1), 1-\rho^2)$

Let's sample from the posterior distribution using a Gibbs sampler assuming $y=(0,0)$ and $\rho=0.8$.

First, we set some of the needed parameters.

```{r}
# set parameters
rho = .8
sigma = sqrt(1 - rho^2)
#observed data
y1 = 0
y2 = 0
```

Let's create a function, `gibbs`, to do the sampling. The function takes:

- `B`: the number of cycles to run.
- `theta`: the vector of initial values.

```{r}
gibbs = function(B, theta) {
  #create matrix to store samples
  theta_sims = matrix(0, nrow = B + 1, ncol = 2)
  theta_sims[1,] = theta
  # run gibbs sampler for B cycles
  for (i in 2:(B+1)) {
    # determine full conditional mean for theta1
    m1 = y1 + rho * (theta[2] - y2)
    # simulate from full conditional distribution for theta1
    theta[1] = rnorm(1, m1, sigma)
    # determine full conditional mean for theta2
    m2 = y2 + rho * (theta[1] - y1)
    # simulate from full conditional distribution for theta1
    theta[2] = rnorm(1, m2, sigma)
    # save sample
    theta_sims[i, ] = theta
  }
  return(theta_sims)
}
```

Let's run an initial chain of 1000 cycles with a starting value $\theta=(-2.5, -2.5)$.

```{r}
# run initial chain
chain1 = gibbs(B = 1000, theta = c(-2.5, -2.5))
```

Let's draw a "sand" plot of our first chain.

```{r}
plot(chain1, pch = ".",
     xlab = expression(theta[1]),
     ylab = expression(theta[2]))
title("Samples from Gibbs sampler")
```

We now run 3 more chains with different starting values.

```{r, echo=TRUE, fig.show='hide'}
chain2 = gibbs(B = 1000, c(-2.5, 2.5))
chain3 = gibbs(B = 1000, c(2.5, -2.5))
chain4 = gibbs(B = 1000, c(2.5, 2.5))
```

We now plot the samples from all 4 chains in one graphic.

```{r}
# chain 1
plot(chain1, pch = ".",
     xlab = expression(theta[1]),
     ylab = expression(theta[2]))
# chains 2-4
points(chain2, pch = ".", col = "orange")
points(chain3, pch = ".", col = "blue")
points(chain4, pch = ".", col = "grey")
# add legend
legend("topleft", pch = 20,
       col = c("black", "orange", "blue", "grey"),
       legend = c("Chain 1", "Chain 2", "Chain 3", "Chain 4"))
title("Samples from Gibbs sampler")
```

We now plot the first 10 cycles of the 4 chains using the `plot_mcmc_path` function from the **bayesutils** package.

`plot_mcmc_path` takes:

- `x`: a matrix with 2 columns or a list of 2 column matrices.
- `ncycles`: the number of cycles to plot.

```{r}
plot_mcmc_path(
  list(chain1, chain2, chain3, chain4),
  ncycles = 10,
  xlim = c(-2.5, 2.5),
  ylim = c(-2.5, 2.5),
  xlab = expression(theta[1]),
  ylab = expression(theta[2]),
  main = "First 10 cycles of each chain"
)
```

# Example: Normal distribution with $\mu$ and $\sigma^2$ unknown and conjugate prior

*Data distribution*

$y_1,\ldots,y_n \mid \mu,\sigma^2 \stackrel{i.i.d.}{\sim} N(\mu,\sigma^2)$. 


*Prior distribution*

$\mu \mid \sigma^2 \sim N(\mu_0, \sigma^2/\kappa_0)$

$\sigma^2 \sim \text{Inv-}\chi^2_{\nu_0}(\sigma_0^2)$

where 

$p(\sigma^2) \propto (\sigma^2)^{-\left(\frac{\nu_0}{2} + 1\right)} \exp\left(-\frac{\nu_0 \sigma_0^2}{2\sigma^2}\right)$.


*Data density*

$p(y \mid \mu,\sigma^2) \propto$

<br>  
<br>  
<br>  
<br>  

$p(\mu,\sigma^2\mid y) \propto$

<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  

Derive the full conditional distributions

<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  


*Full conditional distributions*

$\mu \mid \sigma^2, y \sim N(\mu_n, \tau_n^2)$ with

$$
\mu_n = \frac{\frac{\kappa_0}
{\sigma^2}\mu_0 + \frac{n}{\sigma^2}\bar{y}}{\frac{\kappa_0}
{\sigma^2} + \frac{n}{\sigma^2}}
$$

and

$$
\tau_n^2=\frac{1}{\frac{\kappa_0}{\sigma^2}+\frac{n}{\sigma^2}}=\frac{\sigma^2}{\kappa_0+n}.
$$
$\sigma^2 \mid \mu, y \sim \text{Inv-}\chi^2_{\nu_n}(s_n^2)$

with 

$$
\nu_n = \nu_0 + n + 1
$$

and

$$
s_n^2=\frac{\mu_0 \sigma_0^2+\kappa_0 (\mu-\mu_0 )^2+(n-1) s^2+n(\bar{y}-\mu)^2}{\nu_n}.
$$

**Example:  Midge Data**
---

Grogan and Wirth (1981) provide data on the wing length in millimeters of nine members of a species of midge (small, two-winged flies).  From these nine measurements, we wish to make inference on the population mean $\mu$ and population variance $\sigma^2$.  Create a Gibbs sampler for the parameters $\mu$ and $\sigma^2$.

Studies of other populations suggest that the true mean should be around 1.9 mm with a standard deviation of 0.1.  However, this
population may be different from the others, so we choose $\kappa_0 = \nu_0 = 1$ so that the prior distributions are only weakly centered around these estimates from other populations.

*Data distribution*

$y_1,\ldots,y_n \mid \mu, \sigma^2 \stackrel{i.i.d.}{\sim} N(\mu,\sigma^2)$

*Prior distributions*

$\mu \mid \sigma^2 \sim N(\mu_0, \sigma^2/\kappa_0)$ with $\mu_0=1.9$ and $\kappa_0=1$.

$\sigma^2 \sim  \text{Inv-}\chi^2_{\nu_0}(\sigma_0^2)$
with $\nu_0=1$ and $\sigma_0^2=(0.1)^2$.

We set our seed for reproducible results

```{r}
set.seed(90)
```

We set the parameters of our prior distributions

```{r}
mu0 = 1.9
k0 = 1 #kappa0
nu0 = 1
sigma0 = 0.1
```

We define the number of cycles, the data, the sample size, the sample mean, and the sample standard deviation

```{r}
B = 500000 # number of cycles
# data
y = c(1.64, 1.70, 1.72, 1.74, 1.82, 1.82, 1.82, 1.90, 2.08)
# number of observations
n = length(y)
# sample mean
ybar = mean(y)
# sample standard deviation
s = sd(y)
```

We use the sample mean and standard deviation for our initial values.

```{r}
#Initial value for mu
mu = mean(y)
#Initial value for sigma
sigma = sd(y)
```

We compute the parameters of our full conditional distributions.

```{r}
kn = k0 + n
mun = (k0 * mu0 + n * ybar) / kn
nun = nu0 + n + 1
```

We now implement our Gibbs sampler.

```{r}
sigmasqpost = numeric(B)
mupost = numeric(B)
for (i in 1:B) {
	mu = rnorm(1, mun, sigma/sqrt(kn))

	#parameter for full conditional posterior of sigma^2
	ssqn = (nu0 * sigma0^2 + k0 * (mu - mu0)^2 + (n - 1) * s^2 + n * (ybar - mu)^2)/nun

	# draw from full conditional of sigmasq
	sigmasq = rinvchisq(1, df = nun, scale = ssqn)
	sigma = sqrt(sigmasq)

	# draw from full conditional of mu
	mupost[i] = mu
	sigmasqpost[i] = sigmasq
}
```

In *Bayesian Data Analysis, 3rd edition*, Gelman et al. provide the exact marginal posterior distributions for $\mu$ and $\sigma^2$:

$\mu \mid y \sim t_{\nu_n}(\mu_n, \tau^2_n/\kappa_n)$

and 

$\sigma^2\mid y \sim \text{Inv-}\chi^2_{\nu_n}(\tau_n^2)$,

where

$\nu_n = \nu_0 + n$

and

$\tau_n^2 = (\nu_0\sigma_0^2 + (n - 1)s^2 + \kappa_0 n/\kappa_n  (\bar{y} - \mu_0)^2)/\nu_n$.

Note that this $\nu_n$ and $\tau_n^2$ differ from the $\nu_n$ and $\tau_n^2$ defined for the full conditional distributions.

Let's compare the results from our Gibbs sampler with the true marginal posterior distributions.

We plot the true and approximate marginal posterior density of $\mu$.

```{r}
# define nun and taunsq
vn = nu0 + n
taunsq = (nu0*sigma0^2 + (n - 1) * s^2 + k0 * n/kn * (ybar - mu0)^2)/vn

# plot approximate posterior density for mu
plot(density(mupost), main = "",
     xlab = "mu", xlim = c(1.6, 2))
# plot true posterior density for mu
x = seq(-1, 3, len = 1001)
lines(x, dst(x, df = nun, mean = mun, sigma = sqrt(taunsq/kn)),
      col = "orange")
title("Posterior Density for mu")
legend("topright", legend = c("Gibbs", "true"), col = c("black", "orange"),
	lwd = 1)
```

We plot the true and approximate marginal posterior density of $\sigma^2$.

```{r}
# plot marginal posterior for sigmasq
plot(density(sigmasqpost), main = "",
     xlab = "sigmasq", xlim = c(0, 0.1))
x = seq(0, 0.1, len = 1001)
lines(x, dinvchisq(x, df = vn, scale = taunsq), col = "orange")
title("Posterior Density for sigmasq")
legend("topright",
       legend = c("Gibbs", "True"),
       col = c("black", "orange"),
       lwd = 1)
```

We compute the empirical posterior quantiles for $\mu$.

```{r}
#posterior quantiles for mu
p = c(0.01, 0.10, 0.25, 0.5, 0.75, 0.90, 0.99)
quantile(mupost, prob = p)
```

We compute the true posterior quantiles for $\mu$.

```{r}
qst(p, df = nun, mean = mun, sigma = sqrt(taunsq/kn))
```

We compute the empirical posterior quantiles for $\sigma^2$.

```{r}
quantile(sigmasqpost, prob = p)
```

We compute the true posterior quantiles for $\sigma^2$.

```{r}
qinvchisq(p, df = vn, scale = taunsq)
```

We compute the empirical 95% central posterior interval for $\mu$.

```{r}
quantile(mupost, c(.025, .975))
```

We compute the true 95% central posterior interval for $\mu$.

```{r}
qst(c(0.025, 0.975), df = nun, mean = mun,
    sigma = sqrt(taunsq/kn))
```

We compute the empirical 95% central posterior interval for $\sigma^2$.

```{r}
quantile(sigmasqpost, c(.025, .975))
```

We compute the true 95% central posterior interval for $\sigma^2$.

```{r}
qinvchisq(c(0.025, 0.975), df = vn, scale = taunsq)
```

Lastly, we plot the first 100 cycles of our Gibbs sampler.

```{r}
plot_mcmc_path(cbind(mupost, sigmasqpost), ncycles = 100,
               xlab = expression(mu), ylab = expression(sigma^2))
```

```{r}
#plot cycles of gibbs sampler for 100 iterations
plot_mcmc_path(cbind(mupost, sigmasqpost),
               ncycles = 100, type = "cycle",
               xlab = expression(mu), ylab = expression(sigma^2))
```

# Example:  Multivariate normal with semi-conjugate prior

*Data distribution*

$\mathbf{y}_1,\mathbf{y}_2,…,\mathbf{y}_n \mid \boldsymbol{\mu},\boldsymbol{\Sigma} \stackrel{i.i.d.}{\sim} N(\boldsymbol{\mu},\boldsymbol{\Sigma})$ with each vector having length $d$.

*Prior distributions*

$\boldsymbol{\mu} \sim N(\boldsymbol{\mu}_0, \boldsymbol{\Lambda}_0)$.

$\boldsymbol{\Sigma} \sim \text{Inv-Wishart}_{\nu_0} ( \mathbf{Κ}_0^{-1})$.

$p(\boldsymbol{\Sigma}) \propto |\boldsymbol{\Sigma}|^{-(\nu_0+d+1)/2} \exp\left(-\frac{1}{2} \text{tr}(\mathbf{K}_0 \boldsymbol{\Sigma}^{-1})\right).$

*Data distribution density*

$p(\mathbf{y} \mid \boldsymbol{\mu}, \boldsymbol{\Sigma}) \propto$

<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  

*Prior density for $\boldsymbol{\mu}$*

<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  


$p(\boldsymbol{\mu}, \boldsymbol{\Sigma} \mid \mathbf{y}) \propto$

<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  

Derive the full conditional distributions:

<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  
<br>  


*Full conditional distributions*

$\boldsymbol{\mu} \mid \boldsymbol{\Sigma}, \mathbf{y} \sim N(\boldsymbol{\mu}_n,\boldsymbol{\Lambda}_n)$, where

$$\boldsymbol{\mu}_n=\boldsymbol{\Lambda_n} (\boldsymbol{\Lambda}_0^{-1} \boldsymbol{\mu}_0+n \boldsymbol{\Sigma}^{-1} \bar{\mathbf{y}})$$

and

$$\boldsymbol{\Lambda}_n^{-1}=\boldsymbol{\Lambda}_0^{-1}+n\boldsymbol{\Sigma}^{-1}.$$

$\boldsymbol{\Sigma}\mid \boldsymbol{\mu}, \mathbf{y} \sim   \text{Inv-Wishart}_{\nu_n} (\mathbf{K}_n^{-1 })$, where

$$\nu_n=\mu_0+n,$$

$$\mathbf{K}_n=\mathbf{K}_0+\mathbf{S}_\mu,$$

and

$$
\mathbf{S}_\mu=\sum_{i=1}^n (\mathbf{y}_i-\boldsymbol{\mu}) (\mathbf{y}_i-\boldsymbol{\mu})^T.
$$

A result from linear algebra says that

$$\sum_{i=1}^n (\mathbf{y}_i-\boldsymbol{\mu})\boldsymbol{\Sigma}^{-1} (\mathbf{y}_i-\boldsymbol{\mu})^T = \text{tr}(\mathbf{S}_\mu \boldsymbol{\Sigma^{-1}})$$

where

$$\mathbf{S}_\mu=\sum_{i=1}^n (\mathbf{y}_i-\boldsymbol{\mu}) (\mathbf{y}_i-\boldsymbol{\mu})^T.$$

$\mathbf{S}_\mu$ is the residual sum of squares matrix for the vectors $\mathbf{y}_1,\mathbf{y}_2,\ldots,\mathbf{y}_n$ if the population mean is presumed to be $\boldsymbol{\mu}$.



**Example: Reading Scores**
---

A sample of 22 children are given reading comprehension tests before and after receiving a particular instructional method.  Each student will then have two scores denoting the pre- and post-instructional scores.

Each row of the data set is a two-dimensional vector representing a single case, distributed $N(\boldsymbol{\mu},\boldsymbol{\Sigma})$.


The exam was designed to give an average score around 50 out of 100, so set $\boldsymbol{μ}_0=(50,50)$.

Since the true mean cannot be below 0 or above 100, we want the prior variances to keep $\boldsymbol{\mu}$ inside this range $[0, 100]^2$ with high probability.

We can use

$$
\boldsymbol{\Sigma} =
\begin{bmatrix}
 625 & 312.5 \\
312.5 & 625
\end{bmatrix},
$$

which would mean that the prior probability that the marginal mean is outside $[0, 100]^2$ is only 0.05.

This also accounts for the fact that the two exams measure similar things, so we think the correlation between the means is around 0.5.

Since $E(\text{Inv-Wishart}_{\nu_0} (\mathbf{K}_0^{-1}))=\mathbf{K}_0/(\nu_0-d-1)$, $\nu_0=4$ will loosely center $\mathbf{\Sigma}$ around $\mathbf{K}_0$.

What is the probability that the post-instruction mean is greater than the pre-instruction mean?

What is the probability that the post-instruction test score is greater than the pre-instruction test score for a new student?

Construct a Gibbs sampler to sample from the posterior distribution for $\boldsymbol{\mu}$ and $\boldsymbol{\Sigma}$.


*Data distribution*

$y \mid \boldsymbol{\mu},\boldsymbol{\Sigma}\stackrel{i.i.d.}{\sim} N(\boldsymbol{\mu},\boldsymbol{\Sigma})$.

*Prior distributions*

$\boldsymbol{\mu}\sim N(\boldsymbol{\mu}_0,\boldsymbol{\Lambda}_0)$ with 

$\boldsymbol{\mu}_0=(50,50)$

and

$$
\boldsymbol{\Lambda}_0 =
\begin{bmatrix}
 625 & 312.5 \\
312.5 & 625
\end{bmatrix}.
$$

$\boldsymbol{\Sigma} \sim \text{Inv-Wishart}_{\nu_0} (\mathbf{K}_0^{-1})$ with

$\nu_0=4$

and

$$\mathbf{K}_0=
\begin{bmatrix}
 625 & 312.5 \\
312.5 & 625
\end{bmatrix}.$$

We begin by loading the data.

```{r}
# load data
y = matrix(c(
59, 77, 43, 39, 34, 46, 32, 26, 42, 38, 38, 43, 55, 68,
67, 86, 64, 77, 45, 60, 49, 50, 72, 59, 34, 38, 70, 48,
34, 55, 50, 58, 41, 54, 52, 60, 60, 75, 34, 47, 28, 48,
35, 33), ncol = 2, byrow = TRUE)
# reformat
y = data.frame(pretest = y[,1], posttest = y[,2])
```

We set some generic parameters.

```{r}
n = nrow(y)
d = ncol(y)
```

We set the prior parameters.

```{r}
mu0 = c(50, 50)
nu0 = 4
L0 = K0 = 25^2 * matrix(c(1, .5, .5, 1), nrow = 2)
```

We calculate necessary summary quantities

```{r}
ybar = colMeans(y)
```

We determine some posterior parameters

```{r}
nun = nu0 + n
```

We set the number of cycles.

```{r}
B = 10000
```

We create initial guesses for $\boldsymbol{\mu}$, $\boldsymbol{\Sigma}$, and $\mathbf{S}_\mu$.

```{r}
mu = ybar
Sigma = cov(y)
# initial guess for Smu
Smu = (n - 1) * var(y)
```

We create some matrices/arrays to store our results.

```{r}
mupost = matrix(0, nrow = B, ncol = d)
Sigmapost = array(0, dim = c(d, d, B))
ytildepost = matrix(0, nrow = B, ncol = d)
```

Finally, we execute our Gibbs sampler.

```{r}
for (i in 1:B) {
	#determine full conditional distribution of mean,
	#simulate from distribution
	Ln = solve(solve(L0) + n * solve(Sigma))
	mun = Ln %*% (solve(L0) %*% mu0 + n * solve(Sigma) %*% ybar)
	mu = c((rmvnorm(1, mu = mun, v = Ln)))

	#determine full conditional distribution of Sigma,
	#simulate from distribution
	Smu = (t(y) - mu) %*% t(t(y) - mu)
	Kn = K0 + Smu

	Sigma = rinvwish(1, df = nun, v = Kn)[,,1]

	mupost[i, ] = mu
	Sigmapost[,,i] = Sigma

	ytildepost[i, ] = rmvnorm(1, mu = mu, v = Sigma)
}
```

We estimate the posterior mean of $\boldsymbol{\mu}$.

```{r}
(mean.mupost = apply(mupost, 2, mean))
```

We estimate the posterior variance of $\boldsymbol{\mu}$.

```{r}
(var.mupost = apply(mupost, 2, var))
```


We estimate the posterior mean of $\boldsymbol{\Sigma}$.

```{r}
(mean.Sigmapost = apply(Sigmapost, c(1, 2), mean))
```

We estimate the posterior variance of $\boldsymbol{\Sigma}$.

```{r}
(var.Sigmapost = apply(Sigmapost, c(1, 2), var))
```

We compute some posterior quantiles for $\boldsymbol{\mu}$.

```{r}
apply(mupost, 2, quantile, prob = c(.01, .25, .5, .75, .99))
```


We plot the posterior densities of the pre- and post-test means.

```{r}
dpretest = density(mupost[,1])
dposttest = density(mupost[,2])
plot(dpretest, type = "l", xlim = range(c(dpretest$x, dposttest$x)),
	ylim = range(c(dpretest$y, dposttest$y)), main = "")
lines(dposttest, col = "orange")
legend("topleft", legend = c("mu pretest", "mu postest"),
	col = c("black", "orange"), lwd = c(1, 1))
```

We estimate $P(\mu_{\text{post}} > \mu_{\text{pre}})$.

```{r}
mean(mupost[,2] > mupost[,1])
```

We estimate $P(\tilde{y}_{\text{post}} > \tilde{y}_{\text{pre}})$.

```{r}
mean(ytildepost[,2] > ytildepost[,1])
```

We visualize the joint posterior density of $\boldsymbol{\mu}$.

```{r}
# estimate bivariate density
d2d = MASS::kde2d(mupost[,1], mupost[,2], n = 50)
# plot results
autoimage::pimage(d2d, col = hcl.colors(64),
                  xlab = "mu pretest", ylab = "mu posttest")
contour(d2d, add = TRUE)
points(mupost[1:100,], pch = 20)
```


